---
title: "TimeSeriesVis"
format: html
editor: visual
---


```{r loading-ggplot2, echo=FALSE, attr.warning=FALSE, message=FALSE}
## functional and vectorized single function from 
## https://stackoverflow.com/a/44660688/1585486

using<-function(...) {
  libs <- unlist(list(...))
  req  <- unlist(lapply(libs, require, character.only = TRUE))
  need <- libs[req == FALSE]
  if(length(need) > 0){ 
    install.packages(need)
    lapply(need, require,character.only = TRUE)
  }
}

using("ggplot2", "here")
```


## The problem statement

These examples are re-used from section 2.6.5 of <https://ggplot2-book.org/getting-started#sec-line>.

The dataset called `economics` from the ggplot2 package, has economic data on the US measured over the last 40 years up until 2010. 


Let's first make simple time series plots of the continuous variables un


```{r running-separate-time-series, echo=TRUE}

ggplot(economics, aes(date, unemploy / pop)) +
  geom_line()
ggplot(economics, aes(date, uempmed)) +
  geom_line()
```

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
